---
title: "Overall descriptives"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, error=FALSE, warning=FALSE, message=FALSE}
## Imports ----
# library(lubridate)
library(here)
library(fs)
library(tidyverse)
library(knitr)
source(here("code" , "utils.R"))

## Constants ----
SAVE_DIR <- config::get("private_save_folder")
FIRST_WEEK <- config::get("first_week")
LAST_WEEK <- config::get("last_week")

## Data ----
death_df <- readRDS(paste0(SAVE_DIR, "cleaned_line_data.RDS"))

## Flag the death data ----
death_df <- death_df %>%
    flag_covid_death() %>%
    mutate(non_covid = 1 - covid,
           all_deaths = 1)

flagged_df <- bind_rows(
    ## Excess deaths are only those NOT from COVID
    death_df %>%
        filter(covid == 0) %>%
        flag_any_poisoning_death() %>%
        flag_drug_poisoning() %>%
        flag_nondrug_poisoning() %>%
        flag_alcohol() %>%
        flag_benzos() %>%
        flag_cocaine() %>%
        flag_opioids() %>%
        flag_meth(),
    death_df %>%
        filter(covid == 1)
) %>% 
    filter(dod <= LAST_WEEK)
```
## Total number of deaths before `r LAST_WEEK`
```{r}
NROW(flagged_df)
```

## Subset to CA residents
```{r}
## Subset to CA residents ----
flagged_df <- flagged_df %>% 
    filter(state_name == "CA") %>% 
    select(-state_name)
```

```{r}
NROW(flagged_df)
```

```{r}
sum(death_df$state_name != "CA")
```


## Total number of deaths by sex
```{r}
flagged_df %>% 
    group_by(sex) %>% 
    summarize(n_deaths = n(),
              median_age = median(age, na.rm = TRUE),
              mean_age = mean(age, na.rm = TRUE),
              sd_age = sd(age, na.rm = TRUE),
              min_age = min(age, na.rm = TRUE),
              max_age = max(age, na.rm = TRUE),
              p25 = quantile(age, .25, na.rm = TRUE),
              p75 = quantile(age, .75, na.rm = TRUE)) %>% 
  kable()
```

## Total number of deaths over time

NOTE 2020 only includes up to `r LAST_WEEK`.
```{r}
flagged_df %>% 
    mutate(year = lubridate::year(dod)) %>% 
    count(year) %>% 
  kable()
```

## Total number of poisonings over time

NOTE 2020 only includes up to `r LAST_WEEK`.
```{r}
flagged_df %>% 
    filter(drug_poisoning == 1) %>% 
    mutate(year = lubridate::year(dod)) %>% 
    count(year) %>% 
  kable()
```

## Total number of poisonings by sex
```{r}
flagged_df %>% 
    filter(drug_poisoning == 1) %>% 
    group_by(sex) %>% 
    summarize(n_deaths = n(),
              median_age = median(age, na.rm = TRUE),
              mean_age = mean(age, na.rm = TRUE),
              sd_age = sd(age, na.rm = TRUE),
              min_age = min(age, na.rm = TRUE),
              max_age = max(age, na.rm = TRUE),
              p25 = quantile(age, .25, na.rm = TRUE),
              p75 = quantile(age, .75, na.rm = TRUE)) %>% 
  kable()
```

## Total number of poisonings by sex and year
```{r}
flagged_df %>% 
    filter(drug_poisoning == 1) %>% 
    mutate(year = lubridate::year(dod)) %>% 
    group_by(sex, year) %>% 
    summarize(n_deaths = n(),
              median_age = median(age, na.rm = TRUE),
              mean_age = mean(age, na.rm = TRUE),
              sd_age = sd(age, na.rm = TRUE),
              min_age = min(age, na.rm = TRUE),
              max_age = max(age, na.rm = TRUE),
              p25 = quantile(age, .25, na.rm = TRUE),
              p75 = quantile(age, .75, na.rm = TRUE)) %>% 
  kable()
```

## Total number of poisonings by age
```{r}
flagged_df %>% 
    filter(drug_poisoning == 1) %>% 
    mutate(age_grp = case_when(
      between(age, -Inf, 17) ~ "Under 18",
      between(age, 18, 39)  ~ "18 to 39",
      between(age, 40, 64) ~ "40 to 64",
      between(age, 65, Inf) ~ "65 and up",
      TRUE ~ NA_character_
    )) %>% 
    group_by(age_grp) %>% 
    summarize(n_deaths = n()) %>% 
  kable()
```

## Total number of poisonings by age and year
```{r}
flagged_df %>% 
    filter(drug_poisoning == 1) %>% 
    mutate(age_grp = case_when(
      between(age, -Inf, 17) ~ "Under 18",
      between(age, 18, 24)  ~ "18 to 24",
      between(age, 25, 39)  ~ "25 to 39",
      between(age, 40, 64) ~ "40 to 64",
      between(age, 65, Inf) ~ "65 and up",
      TRUE ~ NA_character_
    ),
    year = lubridate::year(dod)) %>% 
    group_by(age_grp, year) %>% 
    summarize(n_deaths = n()) %>% 
    group_by(year) %>% 
  mutate(prop_year = n_deaths / sum(n_deaths)) %>% 
  group_by(age_grp) %>% 
  mutate(prop_age = n_deaths / sum(n_deaths)) %>% 
ungroup() %>% 
  kable()
```

## Closer look at education and poisonings by year
```{r}
flagged_df %>% 
    filter(drug_poisoning == 1) %>% 
    mutate(year = lubridate::year(dod)) %>% 
    group_by(educ, year) %>% 
    summarize(n_deaths = n(),
              # median_age = median(age, na.rm = TRUE),
              mean_age = mean(age, na.rm = TRUE),
              sd_age = sd(age, na.rm = TRUE),
              # min_age = min(age, na.rm = TRUE),
              # max_age = max(age, na.rm = TRUE),
              # p25 = quantile(age, .25, na.rm = TRUE),
              # p75 = quantile(age, .75, na.rm = TRUE),
              prop_white = mean(race == "White", na.rm = TRUE),
              prop_hispanic = mean(race == "Hispanic", na.rm = TRUE),
              prop_female = mean(sex == "female", na.rm = TRUE)) %>% 
  arrange(year, educ)  %>% 
  ungroup() %>% 
  kable(digits = c(0, 0, 0, 1, 1, 3, 3, 3))
```

## Closer look at education and poisonings overall
```{r}
flagged_df %>% 
    filter(drug_poisoning == 1) %>% 
    mutate(year = lubridate::year(dod)) %>% 
    group_by(educ) %>% 
    summarize(n_deaths = n(),
              # median_age = median(age, na.rm = TRUE),
              mean_age = mean(age, na.rm = TRUE),
              sd_age = sd(age, na.rm = TRUE),
              # min_age = min(age, na.rm = TRUE),
              # max_age = max(age, na.rm = TRUE),
              # p25 = quantile(age, .25, na.rm = TRUE),
              # p75 = quantile(age, .75, na.rm = TRUE),
              prop_white = mean(race == "White", na.rm = TRUE),
              prop_hispanic = mean(race == "Hispanic", na.rm = TRUE),
              prop_female = mean(sex == "female", na.rm = TRUE)) %>% 
  kable(digits = c(0, 0, 1, 1, 3, 3, 3))
```


## Missing race data in poisoning deaths
```{r}
RAW_FILE <- config::get("raw_data_file")

  ## Ingest ----
    raw_data <- readRDS(RAW_FILE)
    
    ## Recode race  ----
    raw_data <- raw_data %>%
        mutate(
            race = case_when(
                ## Being Hispanic trumps all other racial categorization rules
                hispanic.origin == "Y" ~ "Hispanic",
                ## Mixed, other, unknown.
                race.1.code.final == 99 ~ "Unknown", 
                race.1.code.final != race.2.code.final ~ "Other",
                race.1.code.final != race.3.code.final ~ "Other",
                race.2.code.final != race.3.code.final ~ "Other",
                ## Then do single race
                race.1.code.final == 10 ~ "White",
                race.1.code.final == 20 ~ "Black",
                race.1.code.final %in% c(30, 57, 58) ~ "American Indian / Native American",
                race.1.code.final %in% c(41:49, 52:56, 59) ~ "Asian / Pacific Islander",
                race.1.code.final %in% c(51) ~ "Other",
                ## Some rows missing everything
                is.na(race.1.code.final) &
                    is.na(race.2.code.final) & is.na(race.3.code.final) ~ "Completely Missing",
                ## Search on this to make sure it's correct
                TRUE ~ "You missed a code if this label exists"
            )
        )
            
## Clean up a little
    raw_data <- raw_data %>%
        mutate(age_int = as.integer(age.in.years)) %>% 
        select(
            id, 
            race,
            age = age_int, 
            dod = date.of.death,
            state_name = residence.state.province,
            county_name = county.of.residence.geocode.text,
            county_fip = final.county.of.residence.geocode.nchs,
            ucod = final.cause.of.death.icd10,
            starts_with("record.axis.code")
        ) %>% 
        unite(record_all, starts_with("record.axis.code"), sep = " ") %>% 
        mutate(
            ucod = trimws(ucod), 
            record_all = trimws(gsub(pattern = " NA", replacement = "", record_all))) %>% 
        select(-starts_with("record.axis.code"))

flagged_raw <- raw_data %>% 
    flag_any_poisoning_death() %>%
  flag_drug_poisoning() %>% 
    filter(state_name == "CA",
           dod <= LAST_WEEK) 
  
flagged_raw %>% 
    filter(drug_poisoning == 1) %>% 
    mutate(year = lubridate::year(dod)) %>% 
    group_by(race, year) %>% 
    summarize(n_deaths = n(),
              median_age = median(age, na.rm = TRUE),
              mean_age = mean(age, na.rm = TRUE),
              sd_age = sd(age, na.rm = TRUE),
              min_age = min(age, na.rm = TRUE),
              max_age = max(age, na.rm = TRUE),
              p25 = quantile(age, .25, na.rm = TRUE),
              p75 = quantile(age, .75, na.rm = TRUE)) %>% 
  kable()
```

## Missing data in all deaths
```{r}
flagged_raw %>% 
    mutate(year = lubridate::year(dod)) %>% 
    group_by(race, year) %>% 
    summarize(n_deaths = n(),
              median_age = median(age, na.rm = TRUE),
              mean_age = mean(age, na.rm = TRUE),
              sd_age = sd(age, na.rm = TRUE),
              min_age = min(age, na.rm = TRUE),
              max_age = max(age, na.rm = TRUE),
              p25 = quantile(age, .25, na.rm = TRUE),
              p75 = quantile(age, .75, na.rm = TRUE)) %>% 
  kable()
```
